Combine cell type specific analysis

### 2020-10-14 combined_analysis/CTS/combine.Rmd
initial commit

---
title: "CommonMind Consortium: DLPFC"
subtitle: "Cell type specific analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
params:
  upload: TRUE
  cellType: "Olig"
---

@@

rmarkdown::render("combined_analysis/CTS/Olig.Rmd")

# load("geneExpr_logCPM_bMIND.RDATA")
# saveRDS(geneExpr_logCPM_bMIND, file="geneExpr_logCPM_bMIND.RDS")
# saveRDS(geneExpr_logCPM_bMIND_se, file="geneExpr_logCPM_bMIND_se.RDS")

# ml python
# synapse add --parentid syn23234712 geneExpr_logCPM_bMIND.RDS
# synapse add --parentid syn23234712 geneExpr_logCPM_bMIND_se.RDS
--->

```{r knit2synapse, eval= FALSE}
library(synapser)
library(knit2synapse)
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "combined_analysis/combine_all_DLPFC_by_celltype.Rmd",
                                          parentId ="syn23241817",
                                          folderName = 'Analysis of read count data by cohort: GABA',
                                          overwrite=TRUE, 
                                          knitmd=TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}

library(githubr)
parentId = 'syn23241817';
activityName = paste('Analysis of read count data by cohort:', params$cellType)
activityDescription = paste('Analysis of read count data by cohort:', params$cellType)

thisFileName <- paste0(params$cellType, '.Rmd')

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c('png', 'pdf'),
  cache = TRUE)

colorArray = colors()[grep("grey|gray", colors(), invert=TRUE)]

```

```{r read.counts, eval=FALSE}
# Read raw counts from MPP and HBCC
counts.MPP = fread(synGet('syn21867938')$path, data.table=FALSE)
rownames(counts.MPP) = counts.MPP$Geneid
counts.MPP = counts.MPP[,-c(1:6)]

# HBCC - DLFPC
counts.HBCC = fread(synGet('syn21886235')$path, data.table=FALSE)
rownames(counts.HBCC) = counts.HBCC$Geneid
counts.HBCC = counts.HBCC[,-c(1:6)]

# # HBCC - sgACC
counts.HBCC4 = fread(synGet('syn21867866')$path, data.table=FALSE)
rownames(counts.HBCC4) = counts.HBCC4$Geneid
counts.HBCC4 = counts.HBCC4[,-c(1:6)]

geneCountsMerged = cbind(counts.MPP, counts.HBCC, counts.HBCC4)
```

```{r read.cts.expression}
geneExpr.CTS = readRDS(synGet('syn23234833')$path)
geneExpr.CTS.se = readRDS(synGet('syn23234858')$path)

geneExpr = geneExpr.CTS[,params$cellType,]
geneExpr.se = geneExpr.CTS.se[,params$cellType,]

```

### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs = METADATA_QC_DLPFC_ID
metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 19) %>%
  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_dissection:Brain_Region', 
                                      'rnaSeq_isolation:RNA_Isolation_Batch',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned','rnaSeq_report:Transcripts_Detected'))

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn9922992'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY.HBCC = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c(ID = 'SNP_report:Genotyping_Sample_ID'))

ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY.MPP = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'SNP_report:Genotyping_Sample_ID'))

ANCESTRY = rbind(ANCESTRY.HBCC[,colnames(ANCESTRY.MPP)[-2]], ANCESTRY.MPP[,colnames(ANCESTRY.MPP)[-2]])

# Get genotype ids from synapse. 
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
  dplyr::inner_join(ANCESTRY) %>% 
  dplyr::filter(is.na(`SNP_report:Exclude?`))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%      
  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, Region,
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  		 		 dplyr::filter(SampleID %in% colnames(geneExpr), !is.na(SampleID)) %>%
  		 		 dplyr::filter(Region %in% c('DLPFC')) %>%
			 	 dplyr::mutate(Cohort = forcats::fct_recode(Cohort, `MSSM-Penn-Pitt` = 'MSSM-Penn-Pitt', `NIMH-HBCC`='NIMH-HBCC', `NIMH-HBCC`='NIMH-HBCC-4')) %>%
			 	 dplyr::filter(Dx %in% c('Control', "SCZ"))  %>%
         dplyr::filter(ageOfDeath > 17)

# Set Control to baseline
METADATA$Dx = factor(METADATA$Dx, c('Control', 'SCZ'))

ind = METADATA$SampleID[which(METADATA$ReportExclude == 1 | METADATA$GenotypeExclude)]
writeLines(paste('Following',length(ind),'samples are marked exclude'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$PMI)]
writeLines(paste('Following', length(ind), 'counts are missing PMI'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$ageOfDeath)]
writeLines(paste('Following', length(ind), 'counts are missing age of death'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$EV.1)]
writeLines(paste('Following', length(ind), 'counts are missing ancestry information'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

# drop individuals who where sequenced twice here
tab = table(METADATA$Individual_ID)

dropSamples = sapply( names(tab[tab > 1]), function(id){

  idx = which(METADATA$Individual_ID == id)
  df = METADATA[idx,]

  # remove individuals that have less then max reads
  i = which(df$MappedReads < max(df$MappedReads))
  rownames(METADATA)[idx[i]]
  } )
METADATA = METADATA[!(rownames(METADATA) %in% dropSamples),]
```

```{r cellFractionEstimates}
# include estimated cell fractions in METADATA
df_cellFractions = read.table(synGet('syn22333694')$path, row.names=1)

celFrac_ilr = ilr(df_cellFractions)
colnames(celFrac_ilr) = paste0("cellFrac_ilr_", 1:3)

METADATA = merge(METADATA, df_cellFractions, by.x = 'SampleID', by.y='row.names')
METADATA = merge(METADATA, celFrac_ilr, by.x = 'SampleID', by.y='row.names')
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(geneExpr))

geneExpr = geneExpr[,indToRetain]

rownames(METADATA) = METADATA$SampleID
```

```{r create.voom}
vobj.lst = lapply( levels(METADATA$Cohort), function(chrt){

  # get variance within cohort and sex
  # only keep genes with variance > 0
  # v = apply( geneExpr[,idx], 1, var)
  v = lapply( unique(METADATA$Reported_Gender), function(Gnd){
    idx = with(METADATA, (Cohort == chrt) & (Reported_Gender==Gnd))
    isVar = apply( geneExpr[,idx], 1, var) > 0
    names(isVar[isVar])
    })

  # genes with variance in both sexes
  isexpr = intersect(v[[1]], v[[2]])

  # get expressed genes form this cohort
  E = geneExpr[isexpr, METADATA$Cohort == chrt]

  new("EList", list(E = E, weights = 1/geneExpr.se[rownames(E), colnames(E)]))
  })
names(vobj.lst) = levels(METADATA$Cohort)

# only keep genes common to both cohorts
isexpr = intersect(rownames(vobj.lst[[1]]), rownames(vobj.lst[[2]]))

vobj.lst = lapply(vobj.lst, function(vobj){
  vobj[isexpr,]
  })
```

```{r Create.residuals}

### Create residuals
form.lst = list(
  `MSSM-Penn-Pitt`  = ~ Dx*Reported_Gender + RIN +scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + Institution*ageOfDeath,
  `NIMH-HBCC`       = ~ Dx*Reported_Gender + RIN + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + ageOfDeath)

eval_residuals = function( form ){
  # Compute residuals, but add back in Dx*Reported_Gender
  # for form = ~ Dx*Reported_Gender
  f = function(fit){
    residuals(fit) + get_prediction( fit, form)
  }

  residuals.lst = lapply( vobj.lst, function(vobj){

    i = match(colnames(vobj), rownames(METADATA))
    info = METADATA[i,]

    form_use = form.lst[[info$Cohort[1]]]

    # Compute residuals which adding back Diagnosis and Sex effects
    resid.lst = fitVarPartModel(vobj, form_use, info, fxn = f, BPPARAM = SnowParam(18))  
    do.call(rbind, resid.lst)
    })
  names(residuals.lst) = names(vobj.lst)

  residuals.lst
}

resid.add = list( Dx_Sex  = eval_residuals( ~ Dx*Reported_Gender ),
                  Dx      = eval_residuals( ~ Dx ),
                  Sex     = eval_residuals( ~ Reported_Gender ))

residuals.lst = resid.add[['Dx_Sex']]
```

```{r check.correlation, eval=FALSE}
ngenes = 300
C1 = cor(t(residuals.lst[[1]][1:ngenes,]))
C2 = cor(t(residuals.lst[[2]][1:ngenes,]))

df = data.table(A   = as.numeric(C1[lower.tri(C1)]),
                B   = as.numeric(C2[lower.tri(C2)]))

df$density <- get_density(df$A, df$B, n = 80)
ggplot(df, aes(A, B, color=density)) + geom_point() + geom_point(size=.4) + theme_bw(15) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + geom_abline(color="red") + scale_color_viridis() + xlim(-1,1) + ylim(-1,1) + xlab(paste0('Correlation (', names(residuals.lst)[1], ')')) + ylab(paste0('Correlation (', names(residuals.lst)[2], ')'))
```

# Schizophrenia vs controls

### Test marginals for Scz vs controls
```{r differential.expression.simple}
form.lst = list(
  `MSSM-Penn-Pitt`  = ~ Dx + Reported_Gender + RIN +scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + Institution*ageOfDeath,
  `NIMH-HBCC`       = ~ Dx + Reported_Gender + RIN + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + ageOfDeath)

fit_marginal.lst = lapply( vobj.lst, function(vobj){

  i = match(colnames(vobj), rownames(METADATA))
  info = METADATA[i,]

  form_use = form.lst[[info$Cohort[1]]]

  fit = dream(vobj, form_use, info, BPPARAM = SnowParam(18))  
  eBayes( fit )
  })
names(fit_marginal.lst) = names(vobj.lst)

coefTest = c('DxSCZ', 'Reported_GenderMale', 'ageOfDeath')
```

```{r load.genesets, eval=TRUE}
# Load Gene Ontology database 
gs.GO = get_GeneOntology()
gs.MSigDB = get_MSigDB()

# convert from GeneSetCollection to list used by camera and zenith
gs.GO.collection = recodeToList( gs.GO )
gs.MSigDB.collection = recodeToList( gs.MSigDB )
```

```{r run.zenith, eval=TRUE}
res_zenith = lapply( names(fit_marginal.lst), function(key){

  fit = fit_marginal.lst[[key]]

  res.GO = run_zenith(  fit, 
                        coefTest,
                        gs.GO.collection)
  res.GO$Resource = "GO"   
  res.GO$Cohort = key

  res.MSigDB = run_zenith(fit, 
                          coefTest,
                          gs.MSigDB.collection ) 
  res.MSigDB$Resource = "MSigDB"
  res.MSigDB$Cohort = key

  rbind(res.GO, res.MSigDB)
})
names(res_zenith) = names(fit_marginal.lst)

# only autosomes
res_zenith_sex_autosomes = lapply( names(fit_marginal.lst), function(key){

  fit = fit_marginal.lst[[key]]

  # only keep autosomes, for Reported_GenderMale
  tab = coef(fit) %>% getGeneSymbol()
  geneids = rownames(tab)[tab$Chrom %in% 1:22]
  fit = fit[geneids,]

  res.GO = run_zenith(  fit, 
                        'Reported_GenderMale',
                        gs.GO.collection)
  res.GO$Resource = "GO"   
  res.GO$Cohort = key

  res.MSigDB = run_zenith(fit, 
                          'Reported_GenderMale',
                          gs.MSigDB.collection ) 
  res.MSigDB$Resource = "MSigDB"
  res.MSigDB$Cohort = key

  rbind(res.GO, res.MSigDB)
})
names(res_zenith_sex_autosomes) = names(fit_marginal.lst)
```

## Meta-analysis of zenith results
```{r zenith.meta, eval=TRUE}
res_meta_zenith = zenith_meta_analysis( res_zenith )                  
res_meta_zenith.sex_autosome = zenith_meta_analysis( res_zenith_sex_autosomes )
```

```{r get.signatures, fig.width=10, fig.height=5}
# Get DE signatures
df_DE_results = lapply(fit_marginal.lst, function(fit){
  res = lapply(coefTest, function(coef){
    res = topTable(fit, coef, number=Inf) 
    res$Coef = coef
    res %>% getGeneSymbol()  
    })
  names(res) = coefTest
  res
  })
names(df_DE_results) = names(fit_marginal.lst)
```

```{r test.scz, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each cohort
coef = 'DxSCZ'
# ymin = min(sapply(df_DE_results, function(x) min(x[[coef]]$P.Value)))

figList.Scz = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]][[coef]]

  main = paste(key, coef, sep = ': ' )
  plotVolcano( df, df$Symbol[1:10] ) + ggtitle( main ) #+ ylim(0, -log10(ymin))
})

# volcano plots
plot_grid(plotlist=figList.Scz, nrow=1)
```

## Meta-analysis across cohorts
```{r meta.analysis.DE}
# Store results for each coef and cohort
tab.de.lst = lapply(coefTest, function(coef){    
  tab.lst = lapply( df_DE_results, function(lst){
    lst[[coef]]
  })
  names(tab.lst) = names(df_DE_results)
  tab.lst
  })
names(tab.de.lst) = coefTest

# Create list preparing for meta-analysis for SCZ
df_meta = lapply(coefTest, function(coef){
  # Run meta-analysis
  run_meta_analysis( tab.de.lst[[coef]], "FE") %>% getGeneSymbol
})#, mc.cores=6)
names(df_meta) = coefTest
```

```{r concordance.scz, fig.height=6, fig.width=10}
# Concordance analysis
resConcord.SCZ = plot_concordance(tab.de.lst[[coef]], showGenes = df_meta[[coef]]$Symbol[1:20])

main = "Concordance of SCZ"
plot_grid(  resConcord.SCZ$fig_logFC + ggtitle(main),
            resConcord.SCZ$fig_tstat + ggtitle(main), nrow=1)

data.frame(Cohort = names(tab.de.lst[[coef]]),
          "# DE"  = resConcord.SCZ$n.de,
          pi1     = resConcord.SCZ$pi_single,
          check.names=FALSE) %>% 
          mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)

resConcord.SCZ$tab_pi %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.SCZ$tab_corr %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)
```

```{r plot.meta.analysis.SCZ, fig.height=5, fig.width=10}
# volcano plot for meta-analysis
main = paste('Meta-analysis', coef, sep = ': ' )
plotVolcano( df_meta[[coef]], df_meta[[coef]]$Symbol[1:20] ) + ggtitle( main )

data.frame(pi1 = 1 - qvalue(df_meta[[coef]]$P)$pi0,
          `# DE` = sum(df_meta[[coef]]$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for SCZ meta-analysis") %>% kable_styling(full_width=FALSE)

```

## Plot top examples
```{r plot.forrest.dlpfc, fig.width=8, fig.height=20}  
plot_gene = function(residuals.lst, df_meta, geneSymbol, variable, fill=c("#1E90FF", "#FF8D1E"), mode=""){

  ensGene = rownames(df_meta)[which(df_meta$Symbol == geneSymbol)]

  exprResid = lapply( names(residuals.lst), function(key){
      data.frame(Expression = residuals.lst[[key]][ensGene,])
    })
  exprResid = do.call(rbind, exprResid)

  df = merge(exprResid, METADATA, by="row.names")

  if( is.factor(df[[variable]]) || length(unique(df[[variable]])) < 5 ){
    fig = ggplot(df, aes_string(variable, 'Expression')) + geom_violin(aes_string(fill=variable)) + geom_boxplot(width=0.1, fill="grey80") + scale_fill_manual(values = fill)
  }else{    
    fig = ggplot(df, aes_string(variable, 'Expression')) + geom_point() + geom_smooth(method="lm") #+ scale_color_manual(values = fill)
  }

  fig + theme_bw() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + ylab('Expression residuals') + ggtitle(geneSymbol) + facet_wrap(~Cohort)
}

coef = 'DxSCZ'
figList = lapply(df_meta[[coef]]$Symbol[1:10], function(geneSymbol){
  fig1 = plot_forrest( c(tab.de.lst[[coef]], list("Meta-analysis" = df_meta[[coef]])), geneSymbol)

  fig2 = plot_gene(residuals.lst, df_meta[[coef]], geneSymbol, "Dx") + xlab("Diagnosis")

  plot_grid(fig1, fig2, nrow=1, rel_widths=c(.8, 1))
})

plot_grid(plotlist=figList, ncol=1)
```

### Concordance with other datasets
```{r CMC.concordance}
# CMC Analysis 1
CMC1 <- readr::read_tsv(synGet("syn5607603")$path) %>% 
  dplyr::rename(gene_id = genes,
                `CMC 1 t-statistic` = t) %>% 
  dplyr::select(gene_id, `CMC 1 t-statistic`)
# HBCC-DLPFC microarray
CMC_HBCC <- readr::read_tsv(synGet("syn4941553")$path) %>% 
  dplyr::rename(gene_id = genes,
                `HBCC microarray t-statistic` = t) %>% 
  dplyr::select(gene_id, `HBCC microarray t-statistic`)
# Gandal, Science 2018, Table S1
# Use Tab:DGE
# wget https://science.sciencemag.org/highwire/filestream/720110/field_highwire_adjunct_files/1/aat8127_Table_S1.xlsx
gandal <- readxl::read_xlsx( synGet("syn22255217")$path, sheet = "DGE") %>% 
  dplyr::rename(gene_id = ensembl_gene_id,
                `Gandal, SCZ t-statistic` = `SCZ.t.value`,
                `Gandal, BD t-statistic` = `BD.t.value`,
                `Gandal, ASD t-statistic` = `ASD.t.value`) %>%
  dplyr::select(gene_id, `Gandal, SCZ t-statistic`,
                `Gandal, BD t-statistic`, `Gandal, ASD t-statistic`)
# AMP-AD RNASeq reprocessing GRCh37 - https://www.synapse.org/#!Synapse:syn9702085
MSSM <- readr::read_tsv(synGet("syn10526259")$path) %>%
  dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::select(ensembl_gene_id, Tissue, t) %>%
  dplyr::group_by(ensembl_gene_id, Tissue) %>% 
  dplyr::filter(row_number() == 1) %>% 
  tidyr::pivot_wider(names_from = Tissue,
                     values_from = t) %>%
  dplyr::rename(`AMP-AD MSSM FP AD t-statistic` = FP,
                `AMP-AD MSSM IGF AD t-statistic` = IFG,
                `AMP-AD MSSM PHG AD t-statistic` = PHG,
                `AMP-AD MSSM STG AD t-statistic` = STG,
                gene_id = ensembl_gene_id) %>% 
  dplyr::select(gene_id, `AMP-AD MSSM STG AD t-statistic`) 
MSSM <- MSSM[!is.na(MSSM$`AMP-AD MSSM STG AD t-statistic`),]

MAYO <- readr::read_tsv(synGet("syn8468023")$path) %>% 
  dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::select(ensembl_gene_id, `Tissue.ref`, t) %>%
  dplyr::group_by(ensembl_gene_id, `Tissue.ref`) %>% 
  dplyr::filter(row_number() == 1) %>% 
  tidyr::pivot_wider(names_from = `Tissue.ref`,
                     values_from = t) %>% 
  dplyr::rename(`AMP-AD MAYO CBE AD t-statistic` = CBE,
                `AMP-AD MAYO TCX AD t-statistic` = TCX,
                gene_id = ensembl_gene_id) %>% 
  dplyr::select(-`AMP-AD MAYO TCX AD t-statistic`)

ROSMAP <- readr::read_tsv(synGet("syn8456721")$path) %>%
    dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::rename(gene_id = ensembl_gene_id, 
                `AMP-AD ROSMAP DLPFC AD t-statistic` = t) %>% 
  dplyr::select(gene_id, `AMP-AD ROSMAP DLPFC AD t-statistic`)

out_group <- inner_join(MSSM, MAYO) %>% 
  inner_join(., ROSMAP) %>% 
  inner_join(., gandal)

# merge with meta-analysis

coef = 'DxSCZ'
df = df_meta[[coef]]
df_two = data.frame(gene_id = trim_ensembl_ids(rownames(df)),
                `CMC - meta-analysis` = with(df, logFC/se), check.names=FALSE)

df = tab.de.lst[[coef]][[1]]
df.tmp = data.frame(gene_id = trim_ensembl_ids(rownames(df)),
                `t` = df$t, check.names=FALSE)
colnames(df.tmp)[2] = names(tab.de.lst[[coef]])[1]
df_two = df_two %>% inner_join(., df.tmp)

df = tab.de.lst[[coef]][[2]]
df.tmp = data.frame(gene_id = trim_ensembl_ids(rownames(df)),
                `t` = df$t, check.names=FALSE)
colnames(df.tmp)[2] = names(tab.de.lst[[coef]])[2]
df_two = df_two %>% inner_join(., df.tmp)

# Remove transcript Ids
df_two <- df_two %>% 
  tidyr::separate(gene_id, c("gene_id", "transcript")) %>% 
  dplyr::select(-one_of("transcript"))
df_two <- inner_join(df_two, CMC1, by = c("gene_id")) 
df_two <- inner_join(df_two, CMC_HBCC, by = c("gene_id"))
df_two <- inner_join(df_two, out_group, by = c("gene_id")) %>% 
  group_by(gene_id) %>% 
  dplyr::filter(row_number()==1) %>% 
  tibble::column_to_rownames(var = "gene_id")

colnames(df_two) <- gsub(" t-statistic", "", colnames(df_two))

col2 <- colorRampPalette(rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061")))

cor <- cor(df_two)
# clustering based on distance
hcl = hclust(as.dist(1-cor), method='ward.D2')

results <- pvclust(cor, method.hclust='ward.D2')
# reorder based on clustering
reorder = cor[results$hclust$order, results$hclust$order]

# plot clustering
plot(results, hang=-1)
# plot(hcl, hang=-1)
corrplot(reorder, col = col2(200))
```

```{r show.gene_set_test.marginal, eval=TRUE}
show_zenith_results = function( res, coef, Resource, FDR = 0.05){

  i = which((res$Resource == Resource) & (res$Coef == coef))
  df = res[i,]

	# filter by FDR
	df = df[df$FDR <= FDR,]
  df$Coef = c()
  df$se = c()
  df$Resource = c()

	df %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption=paste(Resource, coef), row.names=FALSE) %>% kable_styling(full_width=FALSE)
}

show_zenith_results( res_meta_zenith, "DxSCZ", "GO", FDR=0.05)
show_zenith_results( res_meta_zenith, "DxSCZ", "MSigDB", FDR=0.05)
```

```{r plot.gene_set_test.marginal, eval=TRUE}
 plot_zenith_results = function( res, coef, Resource, FDR = 0.05){

  i = which((res$Resource == Resource) & (res$Coef == coef))
  df = res[i,]

  # filter by FDR
  df = df[df$FDR <= FDR,]
  df$Coef = c()
  df$se = c()
  df$Resource = c()

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$FDR)) * 1.05
  ggplot(df, aes(Set, -log10(FDR))) + geom_bar(stat="identity", fill="navy") + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~FDR)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed")
}

 plot_zenith_results( res_meta_zenith, "DxSCZ", "GO", FDR=0.05)
```

```{r, fig.width=9}
 plot_zenith_results( res_meta_zenith, "DxSCZ", "MSigDB", FDR=0.05)
```

# Sex Analysis
```{r test.Sex, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each chort
coef = 'Reported_GenderMale'
figList.Sex = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]][[coef]]
  main = paste(key, coef, sep = ': ' )
  plotVolcano( df, df$Symbol[1:20] ) + ggtitle( main )
})

# volcano plots
plot_grid(plotlist=figList.Sex, nrow=1)
```

### Autosomes
```{r test.Sex.autosome, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each chort
coef = 'Reported_GenderMale'
figList.Sex = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]][[coef]]
  main = paste(key, coef, sep = ': ' )
  df = df[df$Chrom %in% 1:22,]
  plotVolcano( df, df$Symbol[1:20] ) + ggtitle( main )
})

# volcano plots
plot_grid(plotlist=figList.Sex, nrow=1)
```

## Meta-analysis across cohorts
```{r concordance.Sex, fig.height=6, fig.width=10}
# Concordance analysis
resConcord.Sex = plot_concordance(tab.de.lst[[coef]], showGenes = df_meta[[coef]]$Symbol[1:20])

main = "Concordance of Sex"
plot_grid(  resConcord.Sex$fig_logFC + ggtitle(main),
            resConcord.Sex$fig_tstat + ggtitle(main), nrow=1)

data.frame(Cohort = names(tab.de.lst[[coef]]),
          "# DE"  = resConcord.Sex$n.de,
          pi1     = resConcord.Sex$pi_single,
          check.names=FALSE) %>% 
          mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)

resConcord.Sex$tab_pi  %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.Sex$tab_corr  %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)
```

```{r concordance.Sex.autosome, fig.height=6, fig.width=10}

tab.de.lst.autosome = lapply(tab.de.lst[[coef]], function(x) x[x$Chrom %in% 1:22,])

# Concordance analysis
df = df_meta[[coef]][df_meta[[coef]]$Chrom %in% 1:22,]
resConcord.Sex = plot_concordance(tab.de.lst.autosome, showGenes = df$Symbol[1:20])

main = "Concordance of Sex - autosomes"
plot_grid(  resConcord.Sex$fig_logFC + ggtitle(main),
            resConcord.Sex$fig_tstat + ggtitle(main), nrow=1)

data.frame(Cohort = names(resConcord.Sex$pi_single),
          "# DE"  = resConcord.Sex$n.de,
          pi1     = resConcord.Sex$pi_single,
          check.names=FALSE) %>% 
          mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 within dataset") %>% kable_styling(full_width=FALSE)

resConcord.Sex$tab_pi %>% mutate_if(is.numeric, format, digits=3)  %>% kable(caption="pi1 for replication") %>% kable_styling(full_width=FALSE)
resConcord.Sex$tab_corr  %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="Correlation between t-statistics") %>% kable_styling(full_width=FALSE)
```

>>>>

---
title: "Combine cell type specific analysis"
subtitle: ""
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)"
  affiliation: | 
    Icahn School of Medicine at Mount Sinai, New York
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
---

@@

rmarkdown::render("combined_analysis/CTS/combine.Rmd")

--->

```{r synapse.parameters, include=FALSE, cache=TRUE, eval=FALSE}

library(githubr)
parentId = 'syn23241817';
activityName = "Combine cell type specific analysis"
activityDescription = "Combine cell type specific analysis"

thisFileName <- 'combine.Rmd'

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c('png', 'pdf'),
  cache = TRUE)
```
